---
title: "Bayesian Estimation of parameters for Survival models using the Cox Proportional model"
author: "Daniela Pico Julian Usuga Deivid Zhang"
date: '2022-06-19'
output: pdf_document
classoption:
- twocolumn
editor_options: 
  markdown: 
    wrap: 72
    
---


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(mortDHS)
library(haven)
library(tidyverse)
library(rstanarm)
library(bayesplot)
library(cowplot)
library(huxtable)
color_scheme_set("red")
load(file = "../modelos.RData") # ruta a modelos.rdata
load("~/GitHub/FinalProjectBayesian/.RData")
load("~/GitHub/FinalProjectBayesian/modelos.RData")
```

## Abstract

The main objective of the study was to predict the mortality of people
in African countries. Using data from the DHS and functions from the
**mortDHS** package, functions were created to support the data from any
given country within the set of DHS surveys, a useful model is generated
to make survival predictions and evaluate the effects associated with
covariables and, using the posterior distributions, test significance
and similarities between these covariables, these probabilistic
interpretations can not be done using non-parametric models as
Kaplan-Meier, which was the first model proposed. Specifically, a
discrimination is obtained between the differences in the survival of
each person depending on their sex and country, these predictions are
made from a Bayesian approach using the **rstanarm** package, which uses
the Stan software to make its estimates.



## 1. Introduction

### 1.1. The problem

In our study we were interested in estimating the mortality on African
countries. These countries sometimes lack of a good and reliable
registry of deaths and without these estimating mortality rates can be
difficult, these rates have prime importance on epidemiological or
socio-economic studies.

In our study we used the survey data provided by DHS, these contain a
large number of variables, people were asked about the survival status
of their siblings so we only include sibling data on our study, we
assess siblings survival, expecting to derivate from these the mortality
of the population.

### 1.2. Study Population

Many countries don't have good registries management on government
institutions and thus they don't have a good record of important events
such as population deaths, the Demographic and Health Surveys (DHS)
facilitates multiple datasets containing data collected from
questionnaires performed in households from a large list of countries.

In our study we will consider the Individual Women's data, which consist
of questionnaires performed to women. They were asked about wherever
they had siblings, their survival status, age and death date (in case
they have died).

### 1.3. Data source

The primary data source was the Demographic and Health Surveys (DHS)
Individual Recode (IR) where each row consist of a woman and their
responses on multiple questions. There were multiple columns but we
selected the ones related to siblings and their survival status, these
where:

-   Sibling sex (male or female)
-   Sibling date of birth
-   Sibling survival status (0 = dead, 1 = alive)
-   Sibling date of death
-   Sibling country of residency
-   On particular, we only selected the data from countries Malawi,
    Senegal and Rwanda. But the functions can be extended to any set
    of countries.

From these we calculated the variable **time**, which has a great
importance on survival analysis in general.

## 2. Basic concepts

### 2.1. Survival analysis

Survival analysis is a collection of statistical methods for study the
time that elapses until an event occurs [1]. The name survival is due to
the fact that the applications of this method are mainly study the times
of death, some fields of the application of survival analysis They
include: medicine, epidemiology, economics, among others. An advantage
of models based on survival analysis is that they allow work with
censored data, censoring occurs when you have some information on the
survival time of a patient but the exact time to failure is not known.

There are three functions of interest for survival analysis, the
survival function denoted $S(t)$, the hazard function denoted $h(t)$,
and the hazard function accumulated denoted by $H(t)$, these will be the
amounts of interest for our study. According to the literature [2],
functions are defined as follows:

Let $T$ be a random variable denoting the survival time of a unit or
person. The survival function, which is denoted $S(t)$ gives the
probability that a unit or person will survives beyond some specific
time t. i.e,

$$S(t)= P(T>t)$$

The survival function $S(t)$ is fundamental in AS since for different
values of $t$ provide crucial information of survival data In some
situations it may be more interest to quantify the risk of failure at a
given instant than to estimate survival; a function of interest in
survival analysis that allows do this is the danger function

The hazard function usually denoted as $h(t)$ or $\lambda(t)$ is given
by:

$$ h(t)=\lim_{\Delta t \to 0}\frac{P(t\leq T+\Delta t|T\geq t)}{\Delta t}$$
The numerator represents the conditional probability of the event
occurring in an infinitesimal interval $[t,t+\Delta t]$ (as
$\Delta t\to 0$) given that the unit has survived to $t$ $(T>t)$.
survived until t (T\>t).

The cumulative hazard is defined as:

$$H(t)=\int_{0}^{t}h(s)ds$$

Where $h(s)$ is the hazard function. exist a one to-one relationship
between the hazard function, the cumulative hazard and the probability
of survival, as follows

$$S(t)=exp(-H(t))=exp(-\int_{0}^{t}h(s)ds)$$

### 2.2. Censorship

An observation in a random variable $T$ is said to be right-censored if
all that is known about of $T$ is that it is greater than some value c.
In AS, $T$ refers at the time of occurrence of some particular event and
a case is considered right-censored if it stops observing before for the
event to occur[3]. In our work, only the right caesura is presented,
since at the time of conducting the survey there were people who had not
yet experienced the event of interest (death).

Survival probability can be estimated non-parametrically over temporal
observations (censored and uncensored) using the Kaplan-Meyer method.
other two Common approaches to modeling survival data consist of
modeling the instantaneous rate of the event as a function of time. This
includes the class of models known as proportional hazards regression
models and non-proportional hazards regression models; the second is to
model the time of the event itself. This includes the class of models
known as accelerated time to failure (AFT) models.

## 3. Bayesian approach and why Bayesian

The main reason to go Bayesian in our study was to be able to make
inference, in particular, to make easy to interpretative conclusions on
our parameters and their significance, without the assumptions of
frecuentists approaches.

## 4. Models

Under this modeling framework, it is proposed to implement a of Bayesian
Cox proportional hazard to later make a comparison versus the
frequentest approach and the Kaplan-Meier estimation.

### 4.1. Kaplan--Meier estimator

The Kaplan-Meier estimator, also known as the limit product estimator,
is a nonparametric method for estimating the survival function. survival
function by maximizing the sample likelihood function. Suppose one has k
different failure times $t_1<t_2<...<t_k$, in each time
$t_j(j=1,2,...k)$ there are $n_j$ subjects that are under observation
and at risk of an event of interest.

The K-M estimator is defined as

$$\hat S_{KM}(t)=\prod_{j:t_j\leq t}^{}[1-\frac{d_j}{n_j}]\hspace{2.5mm}$$$$para\hspace{2.5mm}t_1\leq t\leq t_k\hspace{2.5mm}y\hspace{2.5mm}d_j=\#faults$$

### 4.2. Cox proportional hazards model

Under a hazard scale formulation, we model the hazard of the event for
individual $i$ at time $t$ using the following model:

$$h(t|X_i)=h_0(t)exp(X_i\beta)$$

Where $h_0$ is called the hazard base, that is, $h_0(t)$ is the risk
when all $X_i$ variables are 0. $h_0(t)$ characterizes the way in which
hazard changes as a function of survival time, while the second term
characterizes the way the hazard changes as function of the covariates
at the same time guarantees that this is positive, it is also called the
linear predictor.

The formulation of the Cox model in terms of the hazard and the survival
function are given by: From (2) we note that $$S(t;x)=exp(-H(t;x))$$
Where $H(t;\vec{x})$ is the cumulative hazard for a subject with
covariates $\vec{x}=(x_1,x_2,..x_k)$ Assuming survival time is
continuous

$$H(t;\vec{x})=\int_{0}^{t} h(s;\vec{x})ds=\int_{0}^{t} h_0(s)exp(\beta^T\vec{x})ds$$
$$=exp(\beta^T\vec{x})\int_{0}^{t} h_0(s)ds=
exp(\beta^T\vec{x})H_0(T)$$

Cox model in terms of cumulative hazard. In this expression $H_0(t)$ is
the cumulative baseline hazard. This relationship can be thought of as a
baseline cumulative risk measure which is modified according to the
function

From the above relationship, the Cox model can be formulated in terms of
survival:

$$S(t;\vec{x})=exp(-H(t;\vec{x}))=exp(e^{\beta^T\vec{x}})H_0(t)=$$
$$[exp(-H_0(t)]^{e^{\beta^T\vec{x}}}=[S_0(t)]^{e^{\beta^T\vec{x}}}$$

Cox model in terms of survival. In this expression $S_0(t)$ is the
survival baseline.

## 5. Estimation

As we previously seen, the cox proportional hazard model has two parts,
one is the baseline hazard and the second one is the linear model, for
the linear model we only choose to estimate each of the $\beta$ using
normal prios, but for the hazard baseline

### 5.1. Linear Model

From the Cox Proportional-Hazards Model we can extract the following
expression, which we call the linear model.

$$
\mathrm{exp}(\beta \cdot X_i)
$$

Where $\beta = (\beta_0, \beta_1, ... , \beta_n)$ is the vector of
coefficients and $X_i = (1, x_1, ..., x_n)$ is the vector of
covariables.

On our study the will have vector
$X_i= (1,\text{I}_{\text{sex = female}}, \text{I}_{\text{country = Rwanda}}, \text{I}_{\text{country = Senegal}})$,
that is because we have two variables and both of them are factors, the
indicator for the *male* sex and the country *Malawi* are the reference
represented by the intercept $\beta_0$.

For this particular model we set our priors as such:

$$
\beta_0 \sim N(0, 20)
$$

$$
\beta_{i} \sim N(0,2.5) \ \ \ \text{with   i = 1, 2, 3}
$$

### 5.2. Baseline Hazard

The *rstanarm* package allows specifying the risk base through different models, in our study we evaluated 3 types, through an exponential model: for scale parameter $\lambda_i(t) = exp(xi(t))$ , $h_i(t)=\lambda_i(t)$, Weibull model: for scale parameter $\lambda_i(t) = exp(xi(t))$ and shape parameter $\gamma>0$, $h_i(t)=\gamma t^{\gamma -1}\lambda _i(t)$. however, after making a selection of models we opted to work with the M-splines model


In our study we used an M-Spline model to approximate the hazard
baseline function $h_0$, this approximation consists of $\gamma_l$
coefficients multiplied by each component of the spline, function
calculated inside by the package **splines2**.

So the hazard of dying for the individual $i$ at time $t$ is given by:$$
h_i(t) = \mathrm{exp}(\beta \cdot X_i)\ast\sum_{l = 1}^{L}\gamma_l \text{M}_l(t \ | \ k, \delta) 
$$Where $k = \{k_1, ... ,k_J\}$ is a set of knots given by the user, we
observed that a decent number of knots is around 6 (including the 2
boundary knots, $k_1$ located at the earliest entry time and $k_J$,
located at the lastest event), this number of knots prevents that our
model is too overfitting, We will leave the default degree of the
splines $\delta$ at $3$.

The package **rstanarm** handles the calculation and sets a Dirichlet
prior with concentration parameter of 1, ensuring an non-informative
prior.

## 6. Model Selection

To evaluate the performance of the predictions of each model we use a
cross-validation technique. Through the loo() function that is included
in the *rstanarm* package. The way this package divides the data in the
cross validation is leaving a single observation out, the unit that is
systematically omitted determines the predictive task in which the
cross-validation evaluates the performance of the model, the
computational method implemented is Pareto-smoothed importance sampling.

the Pareto- $\hat{k}$ diagnostic estimates how far an individual
leave-one-out distribution is from the full distribution. If leaving out
an observation changes the posterior too much then importance sampling
is not able to give reliable estimate. If $\hat{k}<0.5$, then the
corresponding component of elpd_loo is estimated with high accuracy. If
$0.5< \hat{k}<0.7$ the accuracy is lower, but still OK. if
$\hat{k}>0.7,$ then importance sampling is not able to provide useful
estimate for that component/observation.

The p_loo is called the effective number of parameters and can be
calculated as the difference between the pd_loo and the log posterior
predictive density without cross-validation. In well behaving cases
p_loo$<N$ and p_loo $<p$, where p is the total number of parameters in
the model. p_loo $>N$ or p_loo $>p$ indicates that the model has very
weak predictive capability.

Evaluating the predictive capacity of each model, a k<0.5 was obtained for all the models, indicating a good estimate, as can be seen in tables 4,5,6 of the annexes section, however, the only model that had a p_loo <p, being $p\_loo=8.3$ and $p =12$, was the model that estimates the base hazard as an M-splines as evidenced in table 4, it is also observed that this also has the lowest SE compared to the others, therefore that our work will be developed based on this model

## 7. Results

Table 1 presents the estimation of the Bayesian method described above, 4 chains were used, for each MCMC chain 2000 iterations were carried out, 1000 heating and 1000 sampling, in addition the convergence was evaluated for each parameter of interest, it was obtained by all the methods evaluated (traceplot, effective sample size, acf plots, Rhat), that the parameters converged as observed in table 1, figure 5 and figure 6 respectively.
Table 2 shows the estimation of the frequentist method of the cox model through the Coxph function of the Survival package.

```{=tex}
\begin{table}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
          & mean   & sd    & 10\%   & 50\%   \\ \hline
Intercept & -0.891 & 0.161 & -1.087 & -0.907 \\ \hline
sex2      & -0.235 & 0.069 & -0.327 & -0.234 \\ \hline
countryrw & 0.326  & 0.080 & 0.224  & 0.326  \\ \hline
countryse & -0.326  & 0.095 & -0.446  & -0.327  \\ \hline
\end{tabular}
\caption{parameter estimation bayesian approach}
\end{center}
\end{table}
```
```{=tex}
\begin{table}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
          & coef     & exp(coef) & se(coef) \\ \hline
sex2      & -0.23247 & 0.79257   & 0.06862\\ \hline
countryrw & 0.32528  & 1.38441   & 0.08013 \\ \hline
countryse & -0.33077 & 0.71837   & 0.09340 \\ \hline
\end{tabular}
\caption{parameter estimation frequentist approach}
\end{center}
\end{table}
```


```{r, fig.height=4,fig.cap="\\label{fig:figs} Posterior probabilities distributions"}
mcmc_areas(mod_spline, regex_pars = "Inter*|sex2|country*", prob = 0.90, prob_outer = 0.95)
```
```{r,fig.cap="\\label{fig:figs} Comparison between the KM curve and the Cox Model"}
ps_check(mod_spline)
```


```{r, fig.height=4,fig.cap="\\label{fig:figs} survival function for men and women from Rwanda respectively"}
plot_grid(p1,
          p2,
          ncol = 2)
```

```{r, fig.height=4,fig.cap="\\label{fig:figs} survival function for men and women from Senegal respectively"}
plot_grid(p3,
          p4,
          ncol = 2)
```

```{r, fig.height=4,fig.cap="\\label{fig:figs} survival function for men and women from Malawi respectively"}
plot_grid(p5,
          p6,
          ncol = 2)
```


\newpage
## 8. Interpretations

* According to the estimates obtained, being a woman decreases the logarithm of the relative risk by 0.235 units when the other variables remain constant, belonging to the country of Rwanda increases the logarithm of the relative risk by 0.326 units when the other variables remain constant, while belonging to the country of Senegal the logarithm of the relative risk decreases by 0.326 units when the other variables remain constant, in addition there is statistical evidence to affirm that these parameters are significant, since in figure 4, no parameter takes the value of 0 and therefore these help to explain the probability survival in the regression model
 * In figures 1, 2 and 3, it is observed that for each country the probability of survival is greater for women, in addition the probability of survival decreases if a person is from the country of Rwanda, since it is the country where the curve falls faster , followed by Malawi, in general terms Senegalese women are the group with the highest probability of survival in the three African countries analyzed.
  * Figure 5 illustratively shows the behavior of the posterior distribution, drawing a vertical line at 0 shows that the posterior distribution for the country Senegal and sex 2 (female) takes negative values, indicating the decrease in the logarithm of the function of relative risk, being the strongest effect the one that takes more negative values or, failing that, more positive (those that are farthest from 0), finally, the country Rwanda takes positive values indicating the increase in the logarithm of the relative risk as had been specified above. 

* Figure 6, which compares the survival curve estimated using the Kaplan-Meier frequentist method (stepped curve) with the Cox proportional hazards model (smooth curve), shows a similar behavior up to approximately month 750, in general terms. , makes a similar approximation for both models.

## 9. Conclusions

* In general terms, the estimates obtained on the Cox proportional hazards model by Bayesian estimation and by frequent estimation were very similar, therefore, it is concluded that the Bayesian estimation had a good performance in its estimation, in addition, the benefits of the Bayesian inference for example, the opportunity to make probability statements on parameters, the most natural handling of group-specific parameters, ease of parameter interpretability.
 * Compared to the initial analysis developed by Kene David Nwosu, it was possible to obtain a deeper approach to the probability of survival, where these effects were measured discriminating against gender and country of belonging.
  * The difference obtained at the end of the survival curves in figure 6 may be due to the fact that the curve estimated by the Cox model when taking into account the covariates mentioned above will show a different behavior, it can be thought that this curve is more similar to the true curve and therefore there was an improvement in the estimation of the survival curve.

\newpage

## 10. Annexes

```{=tex}
\begin{table}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
          & mcse   & Rhat    & n\_eff \\ \hline
Intercept & 0.004  & 1.002   & 1432  \\ \hline
sex2      & 0.001  & 1.000   & 3866 \\ \hline
countryrw & 0.001  & 1.001   & 3419 \\ \hline
countryse & 0.002  & 1.001   & 3404 \\ \hline
\end{tabular}
\caption{MCMC diagnostics}
\end{center}
\end{table}
```
```{=tex}
\begin{table}[h!]
\begin{center}
\begin{tabular}{|lll|}
\hline
\multicolumn{1}{|l|}{}          & \multicolumn{1}{l|}{Estimate} & SE    \\ \hline
\multicolumn{1}{|l|}{elpd\_loo} & \multicolumn{1}{l|}{-7089.7}  & 197.5 \\ \hline
\multicolumn{1}{|l|}{p\_loo}    & \multicolumn{1}{l|}{8.3}      & 0.3   \\ \hline
\multicolumn{1}{|l|}{looic}     & \multicolumn{1}{l|}{14179.5}  & 395.1 \\ \hline
\multicolumn{3}{|l|}{Monte Carlo SE of elpd\_loo is 0.0.}               \\ \hline
\multicolumn{3}{|l|}{All Pareto k estimates are good (k \textless 0.5)} \\ \hline
\end{tabular}
\caption{loo splines}
\end{center}
\end{table}
```
```{=tex}
\begin{table}[h!]
\begin{center}
\begin{tabular}{|lll|}
\hline
\multicolumn{1}{|l|}{}          & \multicolumn{1}{l|}{Estimate} & SE    \\ \hline
\multicolumn{1}{|l|}{elpd\_loo} & \multicolumn{1}{l|}{-7429.6}  & 205.2 \\ \hline
\multicolumn{1}{|l|}{p\_loo}    & \multicolumn{1}{l|}{4.2}      & 0.1   \\ \hline
\multicolumn{1}{|l|}{looic}     & \multicolumn{1}{l|}{14859.1}  & 410.4 \\ \hline
\multicolumn{3}{|l|}{Monte Carlo SE of elpd\_loo is 0.0.}               \\ \hline
\multicolumn{3}{|l|}{All Pareto k estimates are good (k \textless 0.5)} \\ \hline
\end{tabular}
\caption{loo exponential}
\end{center}
\end{table}
```
```{=tex}
\begin{table}[h!]
\begin{center}
\begin{tabular}{|lll|}
\hline
\multicolumn{1}{|l|}{}          & \multicolumn{1}{l|}{Estimate} & SE    \\ \hline
\multicolumn{1}{|l|}{elpd\_loo} & \multicolumn{1}{l|}{-7187.3}  & 199.7 \\ \hline
\multicolumn{1}{|l|}{p\_loo}    & \multicolumn{1}{l|}{4.8}      & 0.1  \\ \hline
\multicolumn{1}{|l|}{looic}     & \multicolumn{1}{l|}{14374.7}  & 399.4 \\ \hline
\multicolumn{3}{|l|}{Monte Carlo SE of elpd\_loo is 0.0.}               \\ \hline
\multicolumn{3}{|l|}{All Pareto k estimates are good (k \textless 0.5)} \\ \hline
\end{tabular}
\caption{loo Weibull}
\end{center}
\end{table}
```
```{=tex}
\begin{table}[h!]
\begin{center}
\begin{tabular}{|l|l|l|}
\hline
             & elpd\_diff & se\_diff \\ \hline
mod\_spline  & 0.0        & 0.0      \\ \hline
mod\_weibull & -97.6      & 12.0     \\ \hline
mod\_exp     & -339.8     & 31.2     \\ \hline
\end{tabular}
\caption{loo compare}
\end{center}
\end{table}
```

\newpage
```{r,fig.cap="\\label{fig:figs}Traceplot"}
plot(mod_spline, "trace",pars = "(Intercept)", regex_pars = "country*")
```

```{r, fig.height=4,fig.cap="\\label{fig:figs}ACF"}
plot(mod_spline, "acf", pars = "(Intercept)", regex_pars = "sex2")
```

```{r, fig.height=4,fig.cap="\\label{fig:figs}ACF"}
plot(mod_spline, "acf", pars = "(Intercept)", regex_pars = "country*")
```






\newpage

# References

[1] Allison, PD. (1995). Survival Analysis Using the SAS System: A
practical guide, Cary, NC:SAS Institute Inc., 292 pp.

[2] COX, D. R. Regression Models and Life Tables (with Discussion),
Journal of The Royal Statistical Society, Series B, 34, 187-220, 1972.

[3] ALLISON, P.D. Discrete-Time Methods for the Analysis of Event
Histories, In Sociological Methodology 1982, ed. S. Leinhardt, San
Francisco, CA: Jossey-Bass, 1982.

[4] Ueyama, M. & Yamauchi, F. (2009). Marriage Behavior Response to Prime-Age Adult Mortality: Evidence
from Malawi. Demography, 46(1), 43–63. doi:10.1353/dem.0.0039

[5] Masquelier, B. (2012). Adult Mortality From Sibling Survival Data: A Reappraisal of Selection Biases.
Demography, 50(1), 207–228. doi:10.1007/s13524-012-0149-1